Positivity-Preserving Adaptive Runge-Kutta Methods

### 2020-03-17 latex/overview.tex
Multiple changes to paper: Spelling mistakes Extended section regarding the suitable RKM with short part about existence of b Minor Changes in multiple places
Overview.tex:
Spelling Mistakes
Rewrote some short parts
Changed names to free adaptation and convex adaptation
...

Bib_RK_optim.bib
Added one source,
Removed filenames

Plots:
Multiple corrections, mainly adding legends, changing names, fixing bounding boxes of pdf


Bib_RK_optim.bib, Advection_Decay.pdf und 14 weitere dateien aktualisiert...

When computing the solution of an ODE $u ' = f(t,u) $ using a RKM with $s$ Stages and the Butcher tableau
\begin{align}
\begin{array}{c|c}
c &  A \\
\hline
 & b\\
\end{array}
\end{align}
the stage values $y_1,\cdots,y_s$ are computed according to
\begin{equation}
y_j =  u^n + \dt \sum_{k = 1}^{s} a_{jk} f(t^n + \dt c_k,y_k)  \quad j = 1,\cdots,s
\end{equation}
Based on these values the next solution $u^{n+1}$ is computed by
\begin{equation}
u^{n+1} = u^n + \dt \sum_{j  = 1}^s f(t^n + \dt c_j,y_j) b_j .
\end{equation}
The stage derivatives $f_1,\cdots,f_s$ can be calculated with
\begin{equation}
f_j = f(t^n + \dt c_j,y_j)
\end{equation}
The new solution can now be calculated with
\begin{equation}\label{eq:Combination}
u^{n+1} = u^n + \dt F b
\end{equation}
where $F$ is a matrix containing the stage derivatives $F = \big[f(y_1) , \cdots f(y_s)\big]$.

The new solution $u^{n+1}$ does not comply with positivity for some cases. 
So it might be advantageous to adapt \eqref{eq:Combination} in such a way that $u^{n+1}$ is positive, instead of rejecting the solution and computing a new set of stage values. This is done by adapting the weights.

For the computation of the stage derivatives only the $A$ has to be known. The weights is not needed for this computation. 
Based on this observation we can define $(A,\bullet)$ as the set of RKMs that share the same $A$ and only differ in the weights.
Calculating the stage derivatives can be considered as solving the ODE for all RKM in $(A,\bullet)$ at the same time.
The idea is to choose one $(A,b) \in (A,\bullet)$ by fixing the $b$. 
Because the stage derivatives are already known we can choose a $(A,b)$ that leads to a positive $u^{n+1}$, if $(A,b) \in (A,\bullet) | u^{n+1} \geq 0$ exists.
When altering the weights the properties of a RKM are preserved. 
Especially this approach preserves linear invariants like mass conservation, because it is a regular RKM.

As it can be seen in \eqref{eq:Combination} $u^{n+1}$ is a linear combination of the stage values and the previous solution $u^n$.
Because the coefficients $A$ are already fixed the Order Conditions reduce to linear equations.
This means that finding the weights can be done by solving a linear Programming problem.

@@

\begin{subequations}
\label{eq:Reaction}
\begin{align}
u_1' &= 0.01u_2 + 0.01 u_3 +0.003u_4 - \frac{u_1 u_2}{0.01+u_1} \\ 
u_2' &= \frac{u_1u_2}{0.01+u_1}-0.01 u_2-0.5(1-\exp(-1.21 u_2^2)) u_3 -0.05 u_2 \\ 
u_3' &= 0.5(1-\exp(-1.21u_2^2)) u_3 - 0.01 u_3 -0.02 u_3 \\ 
u_4' &=0.05 u_2 + 0.02 u_3 + 0.003u_4 
\end{align}
\end{subequations}
and the initial conditions
\begin{equation}
u(0) = (8,2,1,4)^T
\end{equation}
When solved with the Cash-Karp (CK5) method and $\dt = 0.005$ the approximated solution contains negative values. This causes qualitatively wrong solutions to the problem. 
In Figure\,\ref{fig:exampleI} the obtained results are plotted with a dashed line. 
At $t=1.905$ $u_1$ gets negative. This leads to a diverging solution.

Now the weights are adapted. The results are also plotted in Figure\,\ref{fig:exampleI}. 
All $u\geq0$ is now fulfilled. A qualitatively correct solution is obtained.

\begin{figure}[h]
    \centering
    \includegraphics[width=0.75\textwidth]{plots/exampleI.pdf}
    \caption{Numerical approximation of the Reaction Problem from \cite{kopecz_comparison_2019}. Computed with ck5 and $\dt = 0.005$. The dashed lines are the reference (Note: Reference might be misleading here because real solution is different) solution without the adaption of the weights. The lower plot are the weights }
    \label{fig:exampleI}
\end{figure}

The used weights $b_1,\cdots,b_5$ are also plotted in Figure\,\ref{fig:exampleI}. 
The used weights are equal to the original weights for $t<1.905$. At $t=1.905$ the weights are first adapted to ensure the positivity of the solution. For $t>2.63$ the original set of weights lead to a positive solution. Because of this, the used weights equal the original weights again.

\section{Details (TODO: better title)}\label{sec:LP}

In the following section, it is explained how the approach can be written as a LP. 
For this we have to consider which weights to choose. 
Generally, any $b \in \R$ can be used. 
A first requirement would be that the new solution is positive.
We can further reduce this space by enforcing the order conditions. 
By incorporating the order conditions, the space of possible weights is limited to a subspace of $\R$. %(Note: not linear because 0 is not in subspace)
Finally, we have to give an objective function to get the optimal weights.

@@

The main goal is to choose a $(A,b)$ in such a way that $u^{n+1}$ resembles the solution of the ODE $u(t_{n+1})$. 
A obvious objective is to use a $(A,b)$ with a high order, but this is not enough.
To get a better understanding for the method we consider the behavior of a simple problem.

@@

\begin{equation}
u'(t) = L u(t) \qquad L = \left[\begin{matrix}- 5 & 1\\5 & -1\end{matrix}\right] 
\end{equation}
and and use the three Stage and third order SSP method (SSP33)
\begin{align}
\begin{array}{c|ccc}
0 &  &  & \\
1 & 1 &  & \\
\frac{1}{2} & \frac{1}{4} & \frac{1}{4} & \\
\hline
 & \frac{1}{6} & \frac{1}{6} & \frac{2}{3}\\
\end{array}
\end{align}
The Matrix $L$ has the eigenvalues $\lambda_1 = -6,\lambda_2=0$. The real stability interval of the SSP33 method includes $[-2.5,0]$. The method is absolut stable for this problem with $\dt = \frac{1}{3}$
The stage derivatives are
\begin{equation}
f(y_1)=\left[\begin{matrix}-5\\5\end{matrix}\right]\qquad f(y_2)=\left[\begin{matrix}5\\-5\end{matrix}\right]\qquad f(y_3)=\left[\begin{matrix}-5\\5\end{matrix}\right]\end{equation}
The values of the next step using the standard weights is
\begin{equation}
u_n=\left[\begin{matrix}- \frac{1}{9}\\\frac{10}{9}\end{matrix}\right]
\end{equation}
The new solution contains negative values. 
Therefor we want to adapt the weights to ensure positifity. 
All weights that comply with the order Condition for the first and second order can be expressed as
\begin{equation}
\tilde b=\left[\begin{matrix}\frac{1}{6}\\\frac{1}{6}\\\frac{2}{3}\end{matrix}\right]+\alpha \left[\begin{matrix}\frac{1}{2}\\\frac{1}{2}\\-1\end{matrix}\right]
\end{equation}
With this we have one degree of freedom for the choice of the weights. This is done by the parameter $\alpha$.
If the general expression for the weights is inserted in \eqref{eq:Combination} the general solution is 
\begin{equation}
u^{1} = u^0 + \dt  \left[f(y_1),f(y_2),f(y_3)\right] b =
\left[\begin{matrix}- \frac{1}{9}\\\frac{10}{9}\end{matrix}\right] +\alpha \left[\begin{matrix}\frac{5}{3}\\- \frac{5}{3}\end{matrix}\right]
\end{equation}
By changing the parameter $\alpha$ the weights $b$ and the solution is altered.
With a suitable choice of $\alpha \in \left[\frac{1}{15},\frac{2}{3}\right]$ any $u$ that complies with mass conservation and positivity can be reached. 
By adding additional constraints on the weights the choice of $\alpha$ can be narrowed down. 
An objective function is also needed to make the choice unique. This should be designed in a way to prefer weights that are close to the original weights.

@@

The order conditions for a RKM are a set of Equations depending on $A$, $c$ and $b$. In these Equations $b$ only appears linearly \cite{hairer_solving_1993}.
If $A$ and $c$ are already known, the order conditions for the Order $p$ are a linear equations system and can be written as $Q b = r$.

@@

\subsection{Ways to formulate the Optimization Problem}

There are two possible ways to write down the adaptation of the weights. The first way is to directly adapt the original weights $b$ by adding a correction $\Delta b$ that ensures positivity.
This is the easiest way to adapt the weights. An advantageous property is that the step size can be reduced after calculating the stage values by using the dense output order conditions.  
The direct adaption is explained in more detail in section\,\ref{sec:direct}.

The second way to formulate a LP is to choose weights as a convex combination of a set of weights. This is more useful if some embedded methods with certain properties are known. 
The convex combination also shows more predictable behavior because the resulting $u^{n+1}_b$ is also a convex combination of the solutions $u^{n+1}_{b_a},u^{n+1}_{b_b},\cdots$ 
This is explained in more detail in section\,\ref{sec:convex}

@@

\begin{figure}
    \centering
    \begin{subfigure}[b]{0.45\textwidth}
        \input{Image_b_direct}
        \caption{Direct Adaptation}
        \label{fig:b_direct}
    \end{subfigure}
    \begin{subfigure}[b]{0.45\textwidth}
        \input{Image_b_convex}
        \caption{Convex Combination}
        \label{fig:b_convex}
    \end{subfigure}
    \caption{Graphical representation of the two different approaches to adapt the weights}\label{fig:b_space}
\end{figure}

\subsection{Direct Adaptation}\label{sec:direct}

When directly adapting the weights new weights that complies with the order conditions is chosen.   
The weights $\tilde{b} = b + \Delta b$ consist of the weights of the original method $b$ and an adaption $\Delta b$. The adaption is $\Delta b \in \mathrm{ker}(Q)$. 
This is drawn in Figure\,\ref{fig:b_direct}. The red line represents the subspace of weights that satisfie the order conditions. Both $b$ and $\tilde{b}$ are in this subspace. 
The LP can be expressed as

@@

\subsubsection{Constraints}
The Direct Adaptation approach leads to two sets of constraints on $b$.
The first set of constraints is the positivity \eqref{eq:direct_pos}.
The second set of constraints are the order conditions in \eqref{eq:direct_Order}.

\subsubsection{Objective Function}
A reasonable property of the objective function would be that for the LP only constrained by the order condition the optimal solution is $\tilde{b} = b$. 
This can be achieved by $argmin(f_{optim}(\tilde{b})) = b$. 
In practice, this also means that there is no need to solve the optimization problem if no constraints are violated because the solution of this problem is already known. This is advantageous for performance reasons.
Large values of b can lead to numerical problems. Because $\sum_{i  = 1}^s b_i = 1$, an increase of $||b||$ leads to a combination of negative and positive values. Because the computation is done with a finite precision (and usually in floating point) large $||b||$ can lead to numerical difficulties.    
By changing the weights the stability function is altered. 
For these reasons $||\tilde{b}-b||$ seems like a good choice. The 1-Norm can be implemented in a  LP using slack variables. We choose \eqref{eq:direct_obj}
as objective function.

@@

If the original step $\dt$ was too large and there is no suitable set of weights, the LP is infeasible. In this case we would like to calculate $u^{n+\theta}$ using the stage values instead of rejecting the step. For this the time step is reduced after calculation the stage values. This can be done using the order conditions for dense output. These can be found in \cite{hairer_solving_1993}.
We introduce $\theta \in (0,1]$. The new step taken has the length $\theta \dt$ whereas $\dt$ is the time step used when calculating the stage values.
To calculate a reduced step the order conditions have to be adapted and a new $b(\theta)$ for the objective function has to be generated. 
When adapting the Order Condition to a new $\theta$ the Order Condition Matrix $Q$ stays unchanged whereas the right hand side $r(\theta)$ depends on $\theta$.

@@

\subsection{Convex combination of existing methods}\label{sec:convex}
For some test problems (e.g. the Heat Equation solved with an implicit method of higher order), large changes occur, even though a norm of $\Delta b$ is used as an objective function. A numerical example is given in Section\,\ref{sec:Numeric_Results}.
It is known that the backward Euler method lead to a positive solution for any $\dt$ \cite{hundsdorfer_numerical_2003}. The backward Euler is still only first Order but shows more exact results than a method of higher Order that is distorted by the adaption of the weights, at least at some steps. 
A possible method to fix this is to generate the new b based on a convex combination of embedded methods. 
The new weights are defined as 
\begin{equation}\label{eq:b_convex_def}
\tilde{b} = \sum_i^d g_i b_i = Bg
\end{equation}
where $B=\left[b_1,\cdots,b_d \right]$ is a matrix containing the embedded methods.  
The LP is defined as
\LPblock{
\begin{subequations}
\begin{align}
 u^{n+1}&=u^n+\dt K B g \geq 0 \label{eq:Convex_pos}  \\
 0 \leq g_i &\leq 1  \forall {i \in \{1, \cdots d \}} \quad \sum_{i=1}^d g_i = 1 \label{eq:Convex_cvx} \\
 \min_{g} &\left(\sum_i^d w_i g_i \right) \label{eq:Convex_obj}
\end{align}
\end{subequations}
}

\subsubsection{Constraints}
At first a set of embedded methods of the RKM $b_1,\cdots,b_d$ is needed. 
If all the used weights complie with the order conditions the new method is also of the same Order.
This is straightforward using
$$ Q b = Q (\chi b_{\alpha} + (1-\chi) b_{\beta}) = \chi Q  b_{\alpha} + (1-\chi) Q b_{\beta} = \chi r + (1-\chi) r = r$$
$b_{\alpha}$ and $b_{\beta}$ are the methods that are combined to $b$ using the parameter $\chi$. The Matrix $Q$ and the vector $r$ are the order conditions as defined in Section\,\ref{sec:OrderCond}.
To ensure that \eqref{eq:b_convex_def} yields to a convex combination the constraints \eqref{eq:Convex_cvx} are enforced for the parameters $g_1,\cdots,g_d$.
The positivity constraint changes to \eqref{eq:Convex_pos}

@@

As a first order method, the weights that corresponds to a series of backward Euler steps are added. These are known to yield a positive result for any step size.
It is also possible to add embedded methods of higher order.
An important property of the embedded methods is that they should show the desired stability characteristics.

@@

\section{TODO: Title} \label{sec:integration}

In the previous sections an algorithm for choosing positifity preserving weights $\tilde{b}$ has been presented. 
In the following section, it is explained how this Algorithm can be integrated into the framework of existing solvers.

\subsection{Choice of Baseline Method}
An important property for the used baseline method is the existence of embedded methods and the degrees of freedom for the b.
As noted in Section\,\ref{sec:OrderCond} the number of stages has to be higher than the order. 
Another important property is the number of degrees of freedom for the choice of the new weights. 
These can be calculated using $s-\mathrm{rank}(Q)$.%$\mathrm{dim}(\mathrm{ker}(Q))$.
The number of degrees of freedom for different methods are shown in table\,\ref{table:DOF_exp} for explicit methods and in table\,\ref{table:DOF_imp} for implicit methods.

\begin{table}[h!]
\centering    %Generated below============ 
\begin{tabular}{|l |c c c c c c |} 
 \hline 
Order &1&2&3&4&5&6 \\ 
 \hline Classical RK4&3&2&0&0& - & -  \\ 
 SSPRK(10,4)&9&8&6&4& - & -  \\ 
 Cash-Karp RK5(4)6&5&4&2&1&0& -  \\ 
 Dormand-Prince RK5(4)7&6&5&3&1&0& -  \\ 
 \hline 
 \end{tabular}
 \caption{Degrees of Freedom for Choice of the weights} %Generated above============ 
 \label{table:DOF_exp}
 \end{table}

 \begin{table}[h!]
\centering   %Generated below============ 
 \begin{tabular}{|l |c c c c c c |} 
 \hline 
Order &1&2&3&4&5&6 \\ 
 \hline Implicit Euler&0& - & - & - & - & -  \\ 
 Im-Euler 2&2&1& - & - & - & -  \\ 
 Im-Euler 3&5&4&2& - & - & -  \\ 
 Im-Euler 4&9&8&6&3& - & -  \\ 
 \hline 
 \end{tabular}
 \caption{Degrees of Freedom for Choice of the weights} %Generated above============ 
 \label{table:DOF_imp}
 \end{table}

For all methods with the number of stages equal to the order of the method, there are no degrees of freedom without reducing the order. 
If the classical RK4 method is used the Order has to be reduced more because the RK4 method does not have embedded methods of order 3.
Contrary to this methods with $s > p$ have degrees of freedom even without reducing the order.

Another important property is the stability region. The adaption of the weights does solve issues with stabilitz when used with a too big $\dt$. Therefore the RK-method has to be stable for large $\dt$.

@@

A third property is the existence of a positive solution. This corresponds to the question if there is an embedded first order method that ensures positivity. This is particularly interesting for implicit methods, because the $\dt$ is not limited by the stability. 
If we have an embedded first Order Method we can ensure that there is always a positive solution, which might be of first Order.

>>>>

When computing the solution of an ODE $u ' = f(t,u) $ using a RKM with $s$ stages and the Butcher tableau
\begin{align}
\begin{array}{c|c}
c &  A \\
\hline
 & b\\
\end{array}
\end{align}
the stage values $y_1,\cdots,y_s$ are computed according to
\begin{equation}
y_j =  u^n + \dt \sum_{k = 1}^{s} a_{jk} f(t^n + \dt c_k,y_k)  \quad j = 1,\cdots,s
\end{equation}
Based on these values the next solution $u^{n+1}$ is computed by
\begin{equation}
u^{n+1} = u^n + \dt \sum_{j  = 1}^s f(t^n + \dt c_j,y_j) b_j .
\end{equation}
The stage derivatives $f_1,\cdots,f_s$ can be calculated with
\begin{equation}
f_j = f(t^n + \dt c_j,y_j)
\end{equation}
The new solution can now be calculated with
\begin{equation}\label{eq:Combination}
u^{n+1} = u^n + \dt F b
\end{equation}
where $F$ is a matrix containing the stage derivatives $F = \big[f(y_1) , \cdots f(y_s)\big]$.

The new solution $u^{n+1}$ does not comply with positivity for some cases. 
So it might be advantageous to adapt \eqref{eq:Combination} in such a way that $u^{n+1}$ is positive, instead of rejecting the solution and computing a new set of stage values. This is done by adapting the weights.
The original weights $b$ are replaced by the adapted weights $\tilde{b}$.
With this adaptation we get
\begin{equation}\label{eq:def_b_tilde}
u^{n+1} = u^n + \dt F \tilde{b}
\end{equation}
form \ref{eq:Combination}. 
This equation is contains $\tilde{b}$ as second unknown alongside $u^{n+1}$. 
First we have to determine the adapted weights to compute the new solution afterwards.
For the computation of the stage derivatives only the $A$ has to be known. The weights are not needed for this computation. 
Based on this observation we can define $(A,\bullet)$ as the set of RKMs that share the same $A$ and only differ in the weights.
Calculating the stage derivatives can be considered as solving the ODE for all RKM in $(A,\bullet)$ at the same time.
The idea is to choose one $(A,\tilde{b}) \in (A,\bullet)$ by fixing the weights. 
Because the stage derivatives are already known, we can choose a $(A,\tilde{b})$, that leads to a positive $u^{n+1}$, if $(A,\tilde{b}) \in (A,\bullet) | u^{n+1} \geq 0$ exists.
When altering the weights, the properties of a RKM are preserved. 
Especially this approach preserves linear invariants like mass conservation, because it is a regular RKM.

As it can be seen in \eqref{eq:Combination} $u^{n+1}$ is a linear combination of the stage values and the previous solution $u^n$.
Because the coefficients in $A$ are already fixed, the Order Conditions reduce to linear equations.
This means that finding the weights can be done by solving a LP.

@@

\begin{subequations}
\label{eq:Reaction}
\begin{align}
u_1' &= 0.01u_2 + 0.01 u_3 +0.003u_4 - \frac{u_1 u_2}{0.01+u_1} \\ 
u_2' &= \frac{u_1u_2}{0.01+u_1}-0.01 u_2-0.5(1-\exp(-1.21 u_2^2)) u_3 -0.05 u_2 \\ 
u_3' &= 0.5(1-\exp(-1.21u_2^2)) u_3 - 0.01 u_3 -0.02 u_3 \\ 
u_4' &=0.05 u_2 + 0.02 u_3 + 0.003u_4 
\end{align}
\end{subequations}
and the initial conditions
\begin{equation}
u(0) = (8,2,1,4)^T
\end{equation}
When solved with the Cash-Karp (CK5) method and $\dt = 0.005$ the approximated solution contains negative values. This causes qualitatively wrong solutions to the problem. 
In Figure\,\ref{fig:exampleI} the obtained results are plotted with a dashed line. 
At $t=1.905$ the value of $u_1$ gets negative. This leads to a diverging solution.

Now the weights are adapted. The results are also plotted in Figure\,\ref{fig:exampleI}. 
The positivity constraint is now fulfilled. A qualitatively correct solution is obtained.

\begin{figure}[h]
    \centering
    \includegraphics[width=0.75\textwidth]{plots/exampleI.pdf}
    \caption{Numerical approximation of the Reaction Problem from \cite{kopecz_comparison_2019}. Computed with ck5 and $\dt = 0.005$. The dashed lines are the approximation obtained without the adaption of the weights. The lower plot are the weights. }
    \label{fig:exampleI}
\end{figure}

The used weights are also plotted in Figure\,\ref{fig:exampleI}. 
The used weights are equal to the original weights for $t<1.905$. At $t=1.905$ the weights are first adapted to ensure the positivity of the solution. For $t>2.63$ the original set of weights lead to a positive solution. Because of this, the used weights equal the original weights again.

\section{Adaptive RKM}\label{sec:LP}

In the following Section,  how the approach can be written as a LP. 
For this we have to consider which weights to choose. 
Generally, any $b \in \R^s$ can be used. 
One requirement would be that the new solution is positive.
We can further reduce this space by enforcing the order conditions. 
By incorporating the order conditions, the space of possible weights is limited to a subset of $\R^s$. %(Note: not linear because 0 is not in subspace)
Finally, we have to state an objective function to get the optimal weights.

@@

The main goal is to choose a $(A,b)$ in such a way that $u^{n+1}$ resembles the solution of the ODE $u(t_{n+1})$. 
An obvious objective is to use a $(A,b)$ with a high order, but this is not enough.
To get a better understanding for the method we consider the behavior of a simple problem.

@@

\begin{equation}
u'(t) = L u(t) \qquad L = \left[\begin{matrix}- 5 & 1\\5 & -1\end{matrix}\right] 
\end{equation}
and use the three stage and third order SSP method (SSP33)
\begin{align}
\begin{array}{c|ccc}
0 &  &  & \\
1 & 1 &  & \\
\frac{1}{2} & \frac{1}{4} & \frac{1}{4} & \\
\hline
 & \frac{1}{6} & \frac{1}{6} & \frac{2}{3}\\
\end{array}
\end{align}
The Matrix $L$ has the eigenvalues $\lambda_1 = -6,\lambda_2=0$. The real stability interval of the SSP33 method includes $[-2.5,0]$. The method is absolut stable for this problem with $\dt = \frac{1}{3}$
The stage derivatives are
\begin{equation}
f(y_1)=\left(\begin{matrix}-5\\5\end{matrix}\right)\qquad f(y_2)=\left(\begin{matrix}5\\-5\end{matrix}\right)\qquad f(y_3)=\left(\begin{matrix}-5\\5\end{matrix}\right)\end{equation}
The values of the next step using the standard weights is
\begin{equation}
u_n=\left(\begin{matrix}- \frac{1}{9}\\\frac{10}{9}\end{matrix}\right)
\end{equation}
The new solution contains negative values. 
Therefore, we want to adapt the weights to ensure positivity. 
All weights, that comply with the order Condition for the first and second order can be expressed as
\begin{equation}
\tilde b=\left(\begin{matrix}\frac{1}{6}\\\frac{1}{6}\\\frac{2}{3}\end{matrix}\right)+\alpha \left(\begin{matrix}\frac{1}{2}\\\frac{1}{2}\\-1\end{matrix}\right).
\end{equation}
We have one degree of freedom for the choice of the weights. This is done by the parameter $\alpha$.
If the general expression for the weights is inserted in \eqref{eq:Combination} the general solution is 
\begin{equation}
u^{1} = u^0 + \dt  \left(f(y_1),f(y_2),f(y_3)\right) \tilde{b} =
\left(\begin{matrix}- \frac{1}{9}\\\frac{10}{9}\end{matrix}\right) +\alpha \left(\begin{matrix}\frac{5}{3}\\- \frac{5}{3}\end{matrix}\right)
\end{equation}
By changing the parameter $\alpha$, the weights and the solution are altered.
With a suitable choice of $\alpha \in \left[\frac{1}{15},\frac{2}{3}\right]$ any $u$ that complies with mass conservation and positivity can be reached. 
By adding additional constraints on the weights, the choice of $\alpha$ can be narrowed down. 
An objective function is also needed to make the choice unique. This should be designed in a way to prefer weights that are close to the original weights.

@@

The order conditions for a RKM are a set of Equations depending on $A$, $c$ and $b$. In these Equations $b$ only appears linearly \cite{hairer_solving_1993}.
If $A$ and $c$ are already known, the order conditions for the order $p$ are a linear equations system and can be written as $Q b = r$.

@@

\subsection{Formulation of the Optimization Problem}

There are two possible ways to write down the adaptation of the weights. The first way is to freely adapt the original weights $b$ by adding a correction $\Delta b$ that ensures positivity.
This is the easiest way to adapt the weights. An advantageous property is that the step size can be reduced after calculating the stage values by using the dense output order conditions.  
The free adaption is explained in more detail in Section\,\ref{sec:direct}.

The second way to formulate a LP is to choose weights as a convex adaptation of a set of weights. This is more useful if some embedded methods with certain properties are known. 
The convex adaptation also shows more predictable behavior because the resulting $u^{n+1}_b$ is also a convex adaptation of the solutions $u^{n+1}_{b_a},u^{n+1}_{b_b},\cdots$ 
This is explained in more detail in Section\,\ref{sec:convex}

@@

\begin{figure}
    \centering
    \begin{subfigure}[b]{0.45\textwidth}
        \input{Image_b_direct}
        \caption{Free adaptation}
        \label{fig:b_direct}
    \end{subfigure}
    \begin{subfigure}[b]{0.45\textwidth}
        \input{Image_b_convex}
        \caption{Convex adaptation}
        \label{fig:b_convex}
    \end{subfigure}
    \caption{Graphical representation of the two different approaches to adapt the weights}\label{fig:b_space}
\end{figure}

\subsection{Free Adaptation}\label{sec:direct}

When freely adapting the weights, new weights that complie with the order conditions is chosen.   
The weights $\tilde{b} = b + \Delta b$ consist of the weights of the original method $b$ and an adaption $\Delta b$. The adaption is $\Delta b \in \mathrm{ker}(Q)$. 
This is drawn in Figure\,\ref{fig:b_direct}. The red line represents the subspace of weights that satisfy the order conditions. Both $b$ and $\tilde{b}$ are in this subspace. 
The LP can be expressed as

@@

\subsubsection{Constraints}
The free adaptation approach leads to two sets of constraints on $b$.
The first set of constraints is the positivity \eqref{eq:direct_pos}.
The second set of constraints are the order conditions in \eqref{eq:direct_Order}.

\subsubsection{Objective Function}
A reasonable property of the LP is that the optimal solution for the problem without positivity constraints is $\tilde{b} = b$. 
This can be achieved by $argmin(f_{optim}(\tilde{b})) = b$. 
In practice, this also means that there is no need to solve the optimization problem if no constraints are violated because the solution of this problem is already known. This is advantageous for performance reasons.
Large values of b can lead to numerical problems. Because $\sum_{i  = 1}^s b_i = 1$, an increase of $||b||$ leads to a combination of negative and positive values. Because the computation is done with a finite precision (and usually in floating point) large $||b||$ can lead to numerical difficulties.    
By changing the weights, the stability function is altered. 
For these reasons $||\tilde{b}-b||$ seems like a good choice. The 1-Norm can be implemented in a  LP using slack variables. We choose \eqref{eq:direct_obj}
as objective function.

@@

If the original step $\dt$ was too large and there is no suitable set of weights, the LP is infeasible. In this case we would like to calculate $u^{n+\theta}$ using the stage values instead of rejecting the step. The time step is reduced after calculation the stage values. This can be done using the order conditions for dense output. These can be found in \cite{hairer_solving_1993}.
We introduce $\theta \in (0,1]$. The new step taken has the length $\theta \dt$ whereas $\dt$ is the time step used when calculating the stage values.
To calculate a reduced step the order conditions have to be adapted and a new $b(\theta)$ for the objective function has to be generated. 
When adapting the Order Condition to a new $\theta$ the Order Condition Matrix $Q$ stays unchanged whereas the right hand side $r(\theta)$ depends on $\theta$.

@@

\subsection{Convex Adaptation of Existing Methods}\label{sec:convex}
For some test problems (e.g. the Heat Equation solved with an implicit method of higher order), large changes occur, even though a norm of $\Delta b$ is used as an objective function. A numerical example is given in Section\,\ref{sec:Numeric_Results}.
It is known that the backward Euler method leads to a positive solution for any $\dt$ \cite{hundsdorfer_numerical_2003}. The backward Euler is still only first Order but shows more exact results than a method of higher Order that is distorted by the adaption of the weights, at least at some steps. 
A possible method to fix this is to generate the new b based on a convex adaptation of embedded methods. 
The new weights are defined as 
\begin{equation}\label{eq:b_convex_def}
\tilde{b} = \sum_i^d g_i b_i = Bg
\end{equation}
where $B=\left[b_1,\cdots,b_d \right]$ is a matrix containing the embedded methods.  
The LP is defined as
\LPblock{
\begin{subequations}
\begin{align}
 u^{n+1}&=u^n+\dt K B g \geq 0 \label{eq:Convex_pos}  \\
 0 \leq g_i &\leq 1  \forall {i \in \{1, \cdots d \}} \quad \sum_{i=1}^d g_i = 1 \label{eq:Convex_cvx} \\
 \min_{g} &\left(\sum_i^d w_i g_i \right) \label{eq:Convex_obj}
\end{align}
\end{subequations}
}

\subsubsection{Constraints}
At first a set of embedded methods of the RKM $b_1,\cdots,b_d$ is needed. 
If all the used weights complie with the order conditions, the new method is also of the same order.
This is straightforward using
$$ Q b = Q (\chi b_{\alpha} + (1-\chi) b_{\beta}) = \chi Q  b_{\alpha} + (1-\chi) Q b_{\beta} = \chi r + (1-\chi) r = r$$
$b_{\alpha}$ and $b_{\beta}$ are the methods that are combined to $b$ using the parameter $\chi$. The Matrix $Q$ and the vector $r$ are the order conditions as defined in Section\,\ref{sec:OrderCond}.
To ensure that \eqref{eq:b_convex_def} yields to a convex adaptation the constraints \eqref{eq:Convex_cvx} are enforced for the parameters $g_1,\cdots,g_d$.
The positivity constraint changes to \eqref{eq:Convex_pos}

@@

As a first order method, the weights that correspond to a series of backward Euler steps are added. These are known to yield a positive result for any step size.
It is also possible to add embedded methods of higher order.
An important property of the embedded methods is that they should show the desired stability characteristics.

@@

\section{Properties of Adaptive RKM} \label{sec:integration}

In the previous Sections an algorithm for choosing positifity preserving weights $\tilde{b}$ has been presented. 
In the following Section, it is explained how this Algorithm can be integrated into the framework of existing solvers.

\subsection{Choice of Baseline Method}
An important property for the used baseline method is the existence of embedded methods and the degrees of freedom for the b. 
As noted in Section\,\ref{sec:OrderCond} the number of stages has to be higher than the order.  
Therefore, explicit RKM are interesting, because an explicit RKM of Order $p$ requires at least $p$ stages. 
Fully implicit RKM can achieve higher orders.  
To fulfill $p<s$ the order has to be reduced drastically. Therefore, these methods are not good candidates for the adaptation. 
As stated in \cite{norsett_attainable_1977} the highest achievable order of a Diagonally implicit RKM is $s+1$. This condition is far closer to the condition required for the adaptation. 
Another important property is the number of degrees of freedom for the choice of the new weights.  
These can be calculated using $s-\mathrm{rank}(Q)$. %$\mathrm{dim}(\mathrm{ker}(Q))$. 
The number of degrees of freedom for different explicit methods are shown in Table\,\ref{table:DOF_exp}.

\begin{table}[h!]
\centering    %Generated below============ 
\begin{tabular}{|l c|c c c c c c |} 
 \hline 
Order & &1&2&3&4&5&6 \\ 
 \hline Classical RK4& s=4&3&2&0&0& - & -  \\ 
 SSPRK(10,4)& s=10&9&8&6&4& - & -  \\ 
 Cash-Karp RK5(4)6& s=6&5&4&2&1&0& -  \\ 
 Dormand-Prince RK5(4)7& s=7&6&5&3&1&0& -  \\ 
 \hline 
 \end{tabular}
 \caption{Degrees of Freedom for Choice of the weights} %Generated above============ 
 \label{table:DOF_exp}
 \end{table}

 \begin{table}[h!]
\centering   %Generated below============ 
\begin{tabular}{|l c|c c c c c c |} 
 \hline 
Order & &1&2&3&4&5&6 \\ 
 \hline Implicit Euler& s=1&0& - & - & - & - & -  \\ 
 Lobatto IIIC4& s=4&3&2&1&0&0&0 \\ 
 Radau IIA3& s=3&2&1&0&0&0& -  \\ 
 SDIRK 54& s=5&4&3&1&0& - & -  \\ 
 TR-BDF2& s=3&2&1& - & - & - & -  \\ 
 Im-Euler 2& s=3&2&1& - & - & - & -  \\ 
 Im-Euler 3& s=6&5&4&2& - & - & -  \\ 
 Im-Euler 4& s=10&9&8&6&3& - & -  \\ 
 \hline 
 \end{tabular}
 \caption{Degrees of Freedom for Choice of the weights} %Generated above============ 
 \label{table:DOF_imp}
 \end{table}

For all methods with the number of stages equal to the order of the method, there are no degrees of freedom without reducing the order.  
If the classical RK4 method is used the Order has to be reduced more because the RK4 method does not have embedded methods of order 3. 
Contrary to this, methods with $s > p$ have degrees of freedom even without reducing the order.  
For Cash-Karp RK5 and Dorman-Prince RK5 the Order has to be reduced to get degrees of freedom even though the number of stages is higher than the order.

In Table\,\ref{table:DOF_imp} same was done for implicit methods. 
We can see that the fully implicit methods Lobatto IIIC4 and Radua IIA3 require a drastic reduction of the order as expected. 
The Diagonally implicit SDIRK 54 method only requires an order reduction of one to get one degree of freedom for the weights.  
The TR-BDF2 method even allows adaptations without reducing the order.  
The Implicit-Euler extrapolation methods also exhibit degrees of freedom without a reduction of the order.

Another important property is the stability region. 
The adaption of the weights does not solve issues with stability when used with a too big $\dt$. Therefore, the RK-method has to be stable for large $\dt$.

@@

A third property is the existence of a positive solution. This corresponds to the question if there is an embedded first order method that ensures positivity. This is particularly interesting for implicit methods, because the $\dt$ is not limited by the stability.  
If we have an embedded first order method, we can ensure that there is always a positive solution, which might be of first Order.  
This is the case for the Im-Euler extrapolation methods, because they contain a chain of BE steps. 
For explicit methods the FE method is always embedded, but a reduction of the step size might be required.
